---
title: "MaxHunter"
output: html_document
date: "2023-06-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(dplyr)
```
Read in the beta matrix via RDS
```{r}
setwd("/nfs/turbo/bakulski1/People/data/ffcw/methyl_data/Feb2022_datafreeze")
betaFileName <- "beta_analytic_freeze1.rds" # file.choose()
startrds <- Sys.time()
betas <- readRDS(betaFileName) # 5.9 GB
rdstime = Sys.time() - startrds; rdstime
# Time difference of 41.2518 secs
```
SKIP THIS CHUNK FOR NOW
Try to reformat the data for downstream efficiency
```{r}
# betas <- as.data.table(betas, keep.rownames = TRUE)
# getwd() # Look before you leap!
# fwrite(betas, file = "beta_matrix.csv")

startfread <- Sys.time()
betas <- fread("/home/lutiffan/beta_matrix.csv")
freadtime = Sys.time() - startfread; freadtime
# It takes over a minute the first time you run fread()
# But then it gets faster?
# Time difference of 28.84002 secs (ratio = 0.7)

# TODO: try fst package

# Transpose
# ttime <- Sys.time()
# tbetas <- transpose(betasTable) 
# Sys.time() - ttime
```
A function to detect local maxima following kernel density estimation of beta values.
```{r}
# https://stackoverflow.com/questions/6836409/finding-local-maxima-and-minima

localMaxima <- function(x, zeroThreshold) {
  x <- ifelse(x < zeroThreshold, 0, x)
  # Use -Inf instead if x is numeric (non-integer)
  #y <- diff(c(-.Machine$integer.max, x)) > 0L
  y <- diff(c(-Inf, x)) > 0L
  
  # rle(y)$lengths
  y <- cumsum(rle(y)$lengths)
  y <- y[seq.int(1L, length(y), 2L)]
  if (x[[1]] == x[[2]]) {
    y <- y[-1]
  }
  y
}
```
Test cases for localMaxima(). Numbers refer to row numbers of beta matrix.
```{r}
# A unimodal probe in the middle of 0 and 1
density206677 <- density(betas[206677,], n = 500, from = 0, to = 1, adjust = 1.5)
maxIndex206677 <- localMaxima(density206677$y, 1/ncol(betas))
maxBetas206677 <- density206677$x[maxIndex206677]
maxFitted206677 <- density206677$y[maxIndex206677]
plot(density206677$x, density206677$y)
points(maxBetas206677, maxFitted206677, col = "red", pch=19, 
         cex=1.25)

# A bimodal probe clustered near 1
density208801 <- density(betas[208801,], n = 500, from = 0, to = 1, adjust = 1.5)
maxIndex208801 <- localMaxima(density208801$y, 1/ncol(betas))
maxBetas208801 <- density208801$x[maxIndex208801]
maxFitted208801 <- density208801$y[maxIndex208801]
plot(density208801$x, density208801$y)
points(maxBetas208801, maxFitted208801, col = "red", pch=19, 
         cex=1.25)

```
A function to wrap localMaxima() and optionally display a plot of detected peaks.
```{r}
maxHunter <- function(probeBetas, betaBreaks = 500, densityAdjust = 1.5, 
                      zeroThreshold, show.plot = FALSE, titleProbeNum = "Needs probe number") {
  densityEst <- density(probeBetas, n = betaBreaks, from = 0, to = 1, 
                        adjust = densityAdjust)
  maxIndex <- localMaxima(densityEst$y, zeroThreshold)
  maxBetas <- densityEst$x[maxIndex] # beta value of peak (from 0 to 1)
  maxFitted <- densityEst$y[maxIndex] # peak value (how high is it)
  
  if (show.plot) {
    plot(densityEst$x, densityEst$y, main = paste0("Kernel Density Est. of Probe in row ", titleProbeNum))
    points(maxBetas, maxFitted, col = "red", pch = 19, cex = 1.25)
    return(list("betas" = maxBetas, "fitted" = maxFitted))
  }
  list("betas" = maxBetas, "fitted" = maxFitted)
  # Do we also want to know about the variance around peaks? Hmm
}
```
Test maxHunter
```{r}
# unimodal
maxHunter(betas[206677,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)
# rough in the middle
maxHunter(betas[410978,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)
# We smooth it out with the adjusted params, but we miss a tiny bump in the middle (TODO)
maxHunter(betas[410978,], zeroThreshold = 0.5, densityAdjust = 2, show.plot = TRUE) 
# bimodal
maxHunter(betas[208801,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)
# bimodal, almost invariant? (TODO). Really messy with default parameters
maxHunter(betas[29932,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)
maxHunter(betas[29932,], zeroThreshold = 0.5, show.plot = TRUE, densityAdjust = 3) # This sort of fixes it.
hist(betas[29932,], breaks = 200, xlim = c(0,1))
mean(betas[29932,] > 0.7)
# unimodal. close to zero but not too close
maxHunter(betas[113220,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)
mean(betas[113220,] < 0.3)
# trimodal with a really short peak close to 1
maxHunter(betas[179171,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)

# We pick up really tiny peaks. Do we care?
# Try messing with values that we squash to zero
maxHunter(betas[429812,], zeroThreshold = 1/ncol(betas), show.plot = TRUE)
maxHunter(betas[429812,], zeroThreshold = 0.5, show.plot = TRUE)

# Density gets weird if most points are close to zero or one
# Try messing with "adjust" parameter
# densityAdjust = 3 seems to do a good job of smoothing graphs for 
# invariant CpG sites
maxHunter(betas[154714,], zeroThreshold = 0.5, densityAdjust = 3, show.plot = TRUE)

maxHunter(betas[429300,], zeroThreshold = 0.5, densityAdjust = 3, show.plot = TRUE)


```
Idea: run maxHunter in a conditional way. Once set of parameters for invariant sites (choose threshold) and again for all the sites in-between.
```{r}
betaSummary <- matrix(nrow = nrow(betas), ncol = 3)
colnames(betaSummary) <- c("hypo", "mid", "hyper")

threshold <- 0.1 # hypomethylation is beta < 0.3, but threshold of 0.3 doesn't do a good job

for (i in 1:nrow(betas)) {
  hypo <- mean(betas[i,] < threshold)
  hyper <- mean(betas[i,] > 1 - threshold)
  mid <- 1 - hypo - hyper
  betaSummary[i,] <- c(hypo, mid, hyper)
}

invariant <- (betaSummary[,1] > 0.90 | betaSummary[,3] > 0.90)

peakSummary <- data.table("numPeaks" = numeric(nrow(betas)), 
                          "peakBetas" = vector(mode = "list", 
                                              length = nrow(betas)),
                          "peakFittedValue" = vector(mode = "list", 
                                              length = nrow(betas)))

set.seed(99)
probeSample <- sample(1:nrow(betas), 20)

# Test on some random probes
for (j in seq_along(probeSample)) {
  i <- probeSample[j]
  
  #browser()
  if (invariant[i]) {
    hist(betas[i,], breaks = 200, xlim = c(0,1), probability = TRUE, 
       main = paste0("Histogram of probe in row ", i, " (invariant)"))
    peaks <- maxHunter(probeBetas = betas[i,], densityAdjust = 3, zeroThreshold = 0.5, show.plot = TRUE, titleProbeNum = i)
  } else {
    hist(betas[i,], breaks = 200, xlim = c(0,1), probability = TRUE, 
       main = paste0("Histogram of probe in row ", i))
    peaks <- maxHunter(probeBetas = betas[i,], zeroThreshold = 0.08, show.plot = TRUE, titleProbeNum = i)
  }
  
  #browser()
  peakSummary[i, numPeaks := length(peaks[[1]])]
  peakSummary[i, peakBetas := list(peaks$betas)]
  peakSummary[i, peakFittedValue := list(peaks$fitted)]
}

# Interesting probes
# 261654: four modes?? set zerothreshold = 0.08
```
Now try running on a bunch to see how long it takes
```{r}
startTime <- Sys.time()
for (i in 1:nrow(betas)) {
  
  #browser()
  if (invariant[i]) {
    # hist(betas[i,], breaks = 200, xlim = c(0,1), probability = TRUE, 
    #    main = paste0("Histogram of probe in row ", i, " (invariant)"))
    peaks <- maxHunter(probeBetas = betas[i,], densityAdjust = 3, 
                      zeroThreshold = 0.5, show.plot = FALSE, titleProbeNum = i)
  } else {
    # hist(betas[i,], breaks = 200, xlim = c(0,1), probability = TRUE, 
    #    main = paste0("Histogram of probe in row ", i))
    peaks <- maxHunter(probeBetas = betas[i,], zeroThreshold = 0.08, show.plot = FALSE, titleProbeNum = i)
  }
  
  #browser()
  peakSummary[i, numPeaks := length(peaks[[1]])]
  peakSummary[i, peakBetas := list(peaks$betas)]
  peakSummary[i, peakFittedValue := list(peaks$fitted)]
}

endTime <- Sys.time() - startTime
endTime # Running all probes should take ~20 minutes
```
Summarize
```{r}
peakCounts <- peakSummary[1:1000] %>% count(numPeaks)
barplot(peakCounts$n ~ peakCounts$numPeaks)

```
Implementation 2 functions
```{r}
# Assume input data is not all slope == 0
localMinMax <- function(fitted, zeroThreshold = 0.01) {
  # Testing: try different datasets
  # fitted <- density4Peak$y; zeroThreshold = 1/100 
  # fitted <- density4PeakInv$y; zeroThreshold = 1/100 
  fitted <- ifelse(fitted < zeroThreshold, 0, fitted)
  
  # Cases:
  # 1. If line increasing at start, plusMinus will be TRUE TRUE...
  # until we reach the first stationary point to the right
  
  # 2. If line decreasing/flat at start, plusMinus will be TRUE FALSE ... 
  # detect stationary point at index == 1
  # When slope is no longer zero, we will incorrectly detect a stationary point
  
  # this code cannot distinguish between negative slope and slope of zero
  plusMinus <- diff(c(-Inf, fitted)) > 0L
  
  # Find indices where the sign of the slope changes
  slopeZero <- cumsum(rle(plusMinus)$lengths); slopeZero
  
  if (length(slopeZero) == 1) {
    return(list("maximaIdx" = slopeZero, "minimaIdx" = c(1, 500)))
  }
  
  # browser()
  
  # 1. First slope sign change is a max
  # 2. First (artificial) slope sign change is a max and is located at index == 1
  # 3. First (artificial) slope sign change is just artificial. Second sign change is an inflection point
  whereMax <- seq.int(1L, length(slopeZero), 2L)
  # version 1
  # whereMin <- -whereMax
  
  # version 2: make these indices positive
  whereMin <- (1:length(slopeZero))[-whereMax] 
  
  # Case 3: Line starts out flat
  if (fitted[[1]] == fitted[[2]]) { 
    # Skip artificial first slope sign change (since we start from -Inf)
    whereMax <- whereMax[-1] 
    
    # Skip inflection point
    # version 1
    # whereMin <- c(-2, whereMin) 
    
    # version 2
    whereMin <- whereMin[-1]
    
    # Maybe not a good idea
    # slopeZero <- slopeZero[-2]
  }
  # # version 1
  # maxes <- slopeZero[whereMax]
  # mins <- slopeZero[whereMin]
  # 
  # return(list("maximaIdx" = maxes, "minimaIdx" = mins))
  
  # version 2
  # return(list("slopeZero" = slopeZero, "maximaIdx" = whereMax,
  # "minimaIdx" = whereMin))
  
  # version 3
  leftMin <- whereMax - 1
  if (leftMin[1] == 0) leftMin[1] = 1
  
  rightMin <- whereMax + 1
  if (rightMin[length(rightMin)] > length(slopeZero)) {
    rightMin[length(rightMin)] = length(slopeZero)
  } 
  
  data.frame("peakIdx" = slopeZero[whereMax], "leftMinIdx" = slopeZero[leftMin], 
             "rightMinIdx" = slopeZero[rightMin])
}
```
Implementation 2 test data
```{r}
# Choose a weird distribution as test case
strangeIdx <- 261654 # 238564
testBetas <- betas[strangeIdx,]
testDensity <- density(testBetas, from = 0, to = 1, n = 500, 
                       bw = bw.SJ(betas[strangeIdx,]), adjust = 2)
```
Run tests
```{r}
testminmax <- localMinMax(testDensity$y); testminmax

# # Plot maxima and minima/inflection points
# hist(betas[strangeIdx,], breaks = 500, xlim = c(0,1), probability = TRUE,
#      main = paste0("Histogram of betas of probe # ", strangeIdx, " (", rownames(betas)[strangeIdx], ")"),
#      xlab = "Beta",
#      ylab = "Probability density")
# lines(testDensity, col = "orchid", lwd = 2, pch = 19)
# points(testDensity$x[testminmax$maximaIdx], testDensity$y[testminmax$maximaIdx], col = "red", pch = 19, cex = 1.25)
# points(testDensity$x[testminmax$minimaIdx], testDensity$y[testminmax$minimaIdx], col = "blue", pch = 19, cex = 1.25)

# Plot maxima and minima/inflection points
hist(betas[strangeIdx,], breaks = 500, xlim = c(0,1), probability = TRUE,
     main = paste0("Histogram of betas of probe # ", strangeIdx, " (", rownames(betas)[strangeIdx], ")"),
     xlab = "Beta",
     ylab = "Probability density")
lines(testDensity, col = "orchid", lwd = 2, pch = 19)
points(testDensity$x[testminmax$slopeZero[testminmax$maximaIdx]], testDensity$y[testminmax$slopeZero[testminmax$maximaIdx]], col = "red", pch = 19, cex = 1.25)
points(testDensity$x[testminmax$slopeZero[testminmax$minimaIdx]], testDensity$y[testminmax$slopeZero[testminmax$minimaIdx]], col = "blue", pch = 19, cex = 1.25)

maximaIdx <- testminmax$slopeZero[testminmax$maximaIdx]; maximaIdx
minimaIdx <- testminmax$slopeZero[testminmax$minimaIdx]; minimaIdx

testminmax$slopeZero[max(testminmax$maximaIdx[1] - 1, 1)]
testminmax$slopeZero[min(testminmax$maximaIdx[1] + 1, 500)]
```
Filtering based on size of peak
```{r}
#### ONLY DO THIS IF THERE IS MORE THAN ONE PEAK ####
if (length(testminmax$maximaIdx) > 1) {
  peaks <- testminmax$maximaIdx
  breaks = testminmax$minimaIdx
  if (breaks[1] != 1) breaks <- c(1, breaks)
  if (breaks[length(breaks)] != 500) breaks <- c(breaks, 500)
  
  cutoff <- 0.01
  # Show peak sample size
  #aboveCutoff <- hist(testBetas, breaks = testDensity$x[breaks], freq = FALSE, plot = TRUE)$count/length(betas[strangeIdx,]) > cutoff
  # Skip plotting
  aboveCutoff <- hist(testBetas, breaks = testDensity$x[breaks], freq = FALSE, plot = TRUE)$count/length(betas[strangeIdx,]) > cutoff
  aboveCutoff
  filtered <- peaks[aboveCutoff]
}

# Plot detected peaks after filtering
hist(betas[strangeIdx,], breaks = 500, xlim = c(0,1), freq = FALSE, 
     main = paste0("Histogram of betas of probe # ", strangeIdx, " (", rownames(betas)[strangeIdx], ")"),
     xlab = "Beta",
     ylab = "Probability density")
lines(testDensity, col = "orchid", lwd = 2, pch = 19)
points(testDensity$x[filtered], testDensity$y[filtered], col = "red", pch = 19, cex = 1.25)
points(testDensity$x[testminmax$minimaIdx], testDensity$y[testminmax$minimaIdx], col = "blue", pch = 19, cex = 1.25)
```
Eliminate short peaks too close to other peaks
```{r}
#### ONLY DO THIS IF THERE IS ANY CROWDING ####
if (sum(diff(testDensity$x[filtered]) < 0.1) > 0) {
  finalPeaks <- filtered
  
  for (i in seq_along(finalPeaks)) {
    currentPeak <- finalPeaks[i]
    if (currentPeak == 0) next # Skip if already deactivated
    
    isNeighbor <- finalPeaks != currentPeak &
                            finalPeaks > currentPeak - 50 & 
                            finalPeaks < currentPeak + 50
    
    neighbors <- finalPeaks[isNeighbor]
    shorter <- density4Peak$y[finalPeaks] < density4Peak$y[currentPeak]
    suppressWarnings(finalPeaks[isNeighbor & shorter] <- 0)
  }
  
  finalPeaks <- finalPeaks[finalPeaks != 0]; finalPeaks
  
  hist(betas[strangeIdx,], breaks = 500, xlim = c(0,1), freq = FALSE, 
     main = paste0("Histogram of betas of probe # ", strangeIdx, " (", rownames(betas)[strangeIdx], ")"),
     xlab = "Beta",
     ylab = "Probability density")
  lines(testDensity, col = "orchid", lwd = 2, pch = 19)
  points(testDensity$x[finalPeaks], testDensity$y[finalPeaks], col = "red", pch = 19, cex = 1.25)
}


```

